---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
---

## Loading and preprocessing the data  

----


* Load the necessary libraries

```{r message=F, warning=F }
if (!require("timeDate")) install.packages("timeDate")
if (!require("ggplot2")) install.packages("ggplot2")

```

* Load the activity file (data)

```{r message=F, warning=F}
if(!file.exists('activity.csv'))
    unzip('activity.zip')
activity <- read.csv("activity.csv")

```

* Transform date from factor into date variable

```{r message=F, warning=F}
activity$date <- as.POSIXct(activity$date)
```

## What is mean total number of steps taken per day?

---

* We ignore NA value and plot the histogram

```{r message=F, warning=F}
daily_steps <- aggregate(activity$steps, by = list(activity$date), sum, na.rm=TRUE) 
names(daily_steps) <- c("Date", "steps")

ggplot(daily_steps,aes(x = steps)) +
    ggtitle("Histogram of daily steps") +
    xlab("Steps") +
    ylab("Frequency") +
	geom_histogram()


```

* We calculate the mean and median of the total number of steps taken per day
```{r message=F, warning=F}

mean(daily_steps$steps, na.rm="TRUE")
median(daily_steps$steps, na.rm="TRUE")

```


## What is the average daily activity pattern?

* Calculate the time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r message=F, warning=F}
avg_steps <- aggregate(activity$steps, by = list(activity$interval), mean, na.rm=TRUE)
names(avg_steps) = c("interval", "mean_steps")
```

* Find the 5-minute interval, on average across all the days in the dataset, which contains the maximum number of steps
```{r message=F, warning=F}
max_avg_steps <- avg_steps[which.max(avg_steps$mean_steps),]
```

* Plot the time series
```{r message=F, warning=F}
ggplot(avg_steps, aes(x = interval, y = mean_steps)) + geom_line() + geom_point(x = max_avg_steps$interval, y=max_avg_steps$mean_steps, color='red',size=5)
```

* The interval with the maximum number of steps
```{r message=F, warning=F}
max_avg_steps$interval
```


## Imputing missing values

**The strategy will be to replace the NA values, with the mean value of the steps of the corresponding interval**

*  Merge so that there is a new column that contains the average of the specific interval of each row

```{r message=F, warning=F}
imputed_steps <- merge( x = activity, y = avg_steps, by = "interval", all.x = TRUE)
```


* Find and filter out the NA values, along with tidying up the needed data

```{r message=F, warning=F}
selected <- which(is.na(imputed_steps$steps))
imputed_steps[selected,"steps"] <- imputed_steps[selected,"mean_steps"]
imputed_steps <- imputed_steps[,1:3]
daily_steps2 <- aggregate(imputed_steps$steps, by = list(imputed_steps$date), sum, na.rm=TRUE)
names(daily_steps2) <- c("Date", "steps")
```

* Plot the histogram of daily steps without NA

```{r message=F, warning=F}
ggplot(daily_steps2,aes(x = steps)) +
    ggtitle("Histogram of daily steps without NA") +
    xlab("Steps") +
	ylab("Frequency") +
    geom_histogram()
```


## Are there differences in activity patterns between weekdays and weekends?

* Create a new factor variable with two levels, weekday and weekend 
```{r message=F, warning=F}
imputed_steps$weekday <- ifelse(isWeekday(imputed_steps$date, wday=1:5),"weekday","weekend")
imputed_steps$weekday <- as.factor(imputed_steps$weekday)
```

* Calculate and create a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r message=F, warning=F}
mean_weekday <- aggregate(imputed_steps$steps, by = list(imputed_steps$weekday, imputed_steps$interval), mean, na.rm = TRUE)
names(mean_weekday) = c("weekday", "interval","mean_steps")
ggplot(mean_weekday, aes(x = interval, y = mean_steps)) + ylab("Number of Steps") + geom_line() + facet_grid(weekday~.)
```


